### 2021-02-17 papers/thesis/full/ch3-disk-slices/disk-slices.tex
All files for thesis committed

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for rotated page geometry
\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage[margin=10pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{bbm}
\usepackage{mleftright}
\usepackage{todonotes}

%\bibliographystyle{apalike}

\graphicspath{ {images/} }

%SetFonts

%SetFonts

\usepackage[square,numbers]{natbib}
\usepackage{url}
%\bibliographystyle{elsarticle-harv}
\bibliographystyle{plain}

\newcommand{\bigO}{\mathcal{O}}
\newcommand{\half}{\frac{1}{2}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\No}{\mathbb{N}_0}
\newcommand{\Ylm}{Y^m_l}
\newcommand{\Ylmfull}{Y^m_l(\theta,\varphi)}
\newcommand{\Plm}{\jac^m_l}
\newcommand{\costheta}{\cos\theta}
\newcommand{\sintheta}{\sin\theta}
\newcommand{\cosphi}{\cos\varphi}
\newcommand{\sinphi}{\sin\varphi}
\newcommand{\eimphi}{e^{im\varphi}}
\newcommand{\alphalm}{\alpha^m_l}
\newcommand{\clm}{c^m_l}
\newcommand{\ctilde}{\tilde{c}^m_l}
\newcommand{\ctildemod}{\tilde{c}^{|m|}_l}
\newcommand{\chat}{\hat{c}^m_l}
\newcommand{\chatmod}{\hat{c}^{|m|}_l}
\newcommand{\ddx}{\frac{\mathrm{d}}{\mathrm{d}x}}
\newcommand{\ddy}{\frac{\mathrm{d}}{\mathrm{d}y}}
\newcommand{\pddx}{\frac{\partial}{\partial x}}
\newcommand{\pddy}{\frac{\partial}{\partial y}}
\newcommand{\pddn}{\frac{\partial}{\partial n}}
\newcommand{\dmdxm}{\frac{\mathrm{d}^m}{\mathrm{d}x^m}}

\newcommand{\Atilde}{\tilde{A}_{l,m}}
\newcommand{\Btilde}{\tilde{B}_{l,m}}
\newcommand{\Dtilde}{\tilde{D}_{l,m}}
\newcommand{\Etilde}{\tilde{E}_{l,m}}
\newcommand{\Ftilde}{\tilde{F}_{l,m}}
\newcommand{\Gtilde}{\tilde{G}_{l,m}}
\newcommand{\Alm}{A_{l,m}}
\newcommand{\Blm}{B_{l,m}}
\newcommand{\Dlm}{D_{l,m}}
\newcommand{\Elm}{E_{l,m}}
\newcommand{\Flm}{F_{l,m}}
\newcommand{\Glm}{G_{l,m}}

\newcommand{\xione}{\xi^{(1)}_{n, \lambda}}
\newcommand{\xitwo}{\xi^{(2)}_{n, \lambda}}
\newcommand{\xithree}{\xi^{(3)}_{n, \lambda}}
\newcommand{\xifour}{\xi^{(4)}_{n, \lambda}}

\newcommand{\bigP}{\mathbb{P}}
\newcommand{\Pl}{\mathbb{P}_l}
\newcommand{\gradP}{T\mathbb{P}}
\newcommand{\gradPl}{T\mathbb{P}_l}
\newcommand{\gradY}{\nabla Y}
\newcommand{\gradYlm}{\nabla Y^m_l}
\newcommand{\gradpY}{\nabla^\perp Y}
\newcommand{\gradpYlm}{\nabla^\perp Y^m_l}

\newcommand{\Dlt}{D^\top_l}

\newcommand{\curlyy}{\bm{\mathcal{Y}}}
\newcommand{\blone}{\beta_{l, 1}}
\newcommand{\blzero}{\beta_{l, 0}}
\newcommand{\blmone}{\beta_{l, -1}}
\newcommand{\chivec}{\bm{\chi}_{1,m_s}}
\newcommand{\cgcoeff}{\mathcal{C}}

\newcommand{\alm}{a_{l,m}}
\newcommand{\blm}{b_{l,m}}
\newcommand{\dlm}{d_{l,m}}
\newcommand{\elm}{e_{l,m}}
\newcommand{\flm}{f_{l,m}}
\newcommand{\glm}{g_{l,m}}
\newcommand{\hlm}{h_{l,m}}
\newcommand{\jlm}{j_{l,m}}
\newcommand{\klm}{k_{l,m}}
\newcommand{\almperp}{a_{l,m}^\perp}
\newcommand{\blmperp}{b_{l,m}^\perp}
\newcommand{\dlmperp}{d_{l,m}^\perp}
\newcommand{\elmperp}{e_{l,m}^\perp}
\newcommand{\flmperp}{f_{l,m}^\perp}
\newcommand{\glmperp}{g_{l,m}^\perp}
\newcommand{\hlmperp}{h_{l,m}^\perp}
\newcommand{\jlmperp}{j_{l,m}^\perp}
\newcommand{\klmperp}{k_{l,m}^\perp}

\newcommand{\unitvec}{\hat{\bm{k}}}

\newcommand{\hdop}{H}
\newcommand{\bighdop}{\mathbb{\hdop}}
\newcommand{\hdopnk}{\hdop_{n,k}}
\newcommand{\hdopnkab}{\hdop_{n,k}^{(a,b)}}
\newcommand{\hdopab}{\hdop^{(a,b)}}
\newcommand{\Wab}{{W^{(a,b)}}}
\newcommand{\hdopmj}{\hdop_{m,j}}
\newcommand{\hdopmjab}{\hdop_{m,j}^{(a,b)}}
\newcommand{\alphaab}{\alpha^{(a,b)}}
\newcommand{\betaab}{\beta^{(a,b)}}
\newcommand{\bighdopab}{\bighdop^{(a,b)}}
\newcommand{\Dnt}{D^\top_n}
\newcommand{\Wii}{W^{(1,1)}}
\newcommand{\hdopii}{\hdop^{(1,1)}}
\newcommand{\bighdopii}{{\mathbb{\hdop}^{(1,1)}}}
\newcommand{\hdopoo}{\hdop^{(0,0)}}
\newcommand{\bighdopoo}{{\mathbb{\hdop}^{(0,0)}}}
\newcommand{\hdopoi}{\hdop^{(0,1)}}
\newcommand{\hdopio}{\hdop^{(1,0)}}
%\newcommand{\jac}{{\tilde P}}
\newcommand{\jac}{{P}}
\newcommand{\genjac}{R}
\newcommand{\genjacnmk}{\genjac_{n-k}}
\newcommand{\genjacmmj}{\genjac_{m-j}}
\newcommand{\genjacw}{w_\genjac}
\newcommand{\jacw}{w_P}
\newcommand{\normgenjac}{\omega_\genjac}
\newcommand{\normjac}{\omega_P}

\newcommand{\dx}{\frac{\partial}{\partial x}}
\newcommand{\dy}{\frac{\partial}{\partial y}}
\newcommand{\laplacewii}{\Delta_W^{(1,1)\to(1,1)}}
\newcommand{\laplacewtt}{\Delta_W^{(2,2)\to(0,0)}}
\newcommand{\laplaceoo}{\Delta^{(0,0)\to(2,2)}}
\newcommand{\biharmonic}{_2\Delta_W^{(2,2)\to(2,2)}}

\newcommand{\element}{\tau}
\newcommand{\refelement}{\hat{\tau}}
\newcommand{\FEset}{\mathcal{T}}
\newcommand{\bigW}{\mathbb{W}}
\newcommand{\bigWab}{\mathbb{W}^{(a,b)}}
\newcommand{\bigWii}{{\mathbb{W}^{(1,1)}}}
\newcommand{\bigQ}{\mathbb{Q}}
\newcommand{\bigQab}{\bigQ^{(a,b)}}

\newcommand{\hdopnkabc}{\hdop_{n,k}^{(a,b,c)}}
\newcommand{\hdopabc}{\hdop^{(a,b,c)}}
\newcommand{\Wabc}{{W^{(a,b,c)}}}
\newcommand{\hdopmjabc}{\hdop_{m,j}^{(a,b,c)}}
\newcommand{\alphaabc}{\alpha^{(a,b,c)}}
\newcommand{\betaabc}{\beta^{(a,b,c)}}
\newcommand{\bighdopabc}{\bighdop^{(a,b,c)}}
\newcommand{\Wiii}{W^{(1,1,1)}}
\newcommand{\hdopiii}{\hdop^{(1,1,1)}}
\newcommand{\bighdopiii}{{\mathbb{\hdop}^{(1,1,1)}}}
\newcommand{\hdopooo}{\hdop^{(0,0,0)}}
\newcommand{\bighdopooo}{{\mathbb{\hdop}^{(0,0,0)}}}
\newcommand{\hdopooi}{\hdop^{(0,0,1)}}
\newcommand{\hdopiio}{\hdop^{(1,1,0)}}
\newcommand{\laplacewiii}{\Delta_W^{(1,1,1)\to(1,1,1)}}
\newcommand{\laplacewttt}{\Delta_W^{(2,2,2)\to(0,0,0)}}
\newcommand{\laplaceooo}{\Delta^{(0,0,0)\to(2,2,2)}}
\newcommand{\biharmonictwo}{_2\Delta_W^{(2,2,2)\to(2,2,2)}}
\newcommand{\bigWiii}{{\mathbb{W}^{(1,1,1)}}}
\newcommand{\bigWabc}{\mathbb{W}^{(a,b,c)}}

\newcommand{\alphaabcd}{\alpha^{(a,b,c,d)}}
\newcommand{\betaabcd}{\beta^{(a,b,c,d)}}
\newcommand{\hdopnkabcd}{\hdop_{n,k}^{(a,b,c,d)}}
\newcommand{\hdopabcd}{\hdop^{(a,b,c,d)}}
\newcommand{\Wabcd}{{W^{(a,b,c,d)}}}
\newcommand{\hdopmjabcd}{\hdop_{m,j}^{(a,b,c,d)}}
\newcommand{\bighdopabcd}{\bighdop^{(a,b,c,d)}}
\newcommand{\bigWabcd}{\mathbb{W}^{(a,b,c,d)}}

\newcommand{\genjact}{\tilde{\genjac}}
\newcommand{\genjactnmk}{\genjact_{n-k}}
\newcommand{\genjactmmj}{\genjact_{m-j}}
\newcommand{\genjactw}{w_{\genjact}}
\newcommand{\normgenjact}{\omega_{\genjact}}

\usepackage{amsthm}

\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma} 
\newtheorem{theorem}{Theorem} 
\newtheorem{definition}{Definition}
\newtheorem{condition}{Condition}
\newtheorem{algorithm}{Algorithm}
\newtheorem{corollary}{Corollary}

\def\condref#1{Condition~\ref{cond:#1}}

\include{somacros}

%\usepackage{media9}
%\graphicspath{ {../sphere/plots} }

\title{Sparse spectral and $p$-finite element methods for partial differential equations on disk slices and trapeziums}
\author{Sheehan Olver, Ben Snowball\thanks{Department of Mathematics, Imperial College, London, England}}
%\date{}							% Activate to display a given date or no date

\begin{document}

\maketitle

\begin{abstract}
Sparse spectral methods for solving partial differential equations have been derived in recent years using hierarchies of classical orthogonal polynomials on intervals, disks, and triangles. In this work we extend this methodology to a hierarchy of non-classical orthogonal polynomials on disk slices and trapeziums. This builds on the observation that sparsity is guaranteed due to the boundary being defined by an algebraic curve, and that the entries of partial differential operators can be determined using formulae in terms of (non-classical) univariate orthogonal polynomials. We apply the framework to solving the Poisson, variable coefficient Helmholtz, and Biharmonic equations. In this paper we focus on constant Dirichlet boundary conditions, as well as zero Dirichlet and Neumann boundary conditions, with other types of boundary conditions requiring future work.
 \end{abstract}

%
\section{Introduction}

This paper develops sparse spectral methods for solving linear partial differential equations on a special class of geometries that includes disk slices and trapeziums.  
More precisely, we consider the solution of partial differential equations on the domain
\begin{align*}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align*}
where  either of the following conditions hold:

\begin{condition}\label{cond:trap}
 \(\rho\) is a degree 1 polynomial.
 \end{condition}

 \begin{condition}\label{cond:disk}
\(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\).
\end{condition}

For simplicity of presentation we focus on the disk-slice, where $\rho(x) = \sqrt{1-x^2}$, $(\alpha,\beta) \subset (0, 1)$, and $(\gamma, \delta)  = (-1,1)$, and discuss an extension to other geometries in the appendix (including the half-disk and trapeziums).

We show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that  generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations \cite{olver2013fast} and its analogue on the disk \cite{vasil2016tensor} and triangle \cite{olver2018recurrence,olver2019triangle}.  On the disk-slice the family of weights we consider are of the form
$$
\Wabc(x,y) = (\beta - x)^a \: (x - \alpha)^b \: (1-x^2-y^2)^c, \qfor \alpha \leq x \leq \beta,\quad -\rho(x) \leq y \leq \rho(x).
$$
The corresponding OPs denoted $\hdopnkabc(x,y)$, where $n$ denotes the polynomial degree, and $0 \le k \le n$. We define these to be orthogonalised lexicographically, that is,
$$
\hdopnkabc(x,y) = C_{n,k} x^{n-k} y^k + (\hbox{lower order terms})
$$
where $C_{n,k} \neq 0$ and ``lower order terms''   includes degree $n$ polynomials of the form $x^{n-j} y^j$ where $j < k$. The precise normalization arises from their definition in terms of one-dimensional OPs in Definition~\ref{def:OPconstruction}.

@@

Here is an overview of the paper:

\noindent \secref{OPs}: We present our procedure to gain a (two-parameter) family of 2D orthogonal polynomials (OPs) on the disk-slice domain, by combining 1D OPs on the interval, to form 2D OPs on the disk.

\noindent\secref{PDOs}: We demonstrate that these families will lead to sparse operators, including Jacobi operators representing multiplication by $x$ and $y$, and partial differential operators. We present a method involving use of the Christoffel--Darboux formula \cite[18.2.2]{DLMF} to obtain the recurrence coefficients for the non-classical 1D OPs, allowing us to exactly use the 1D quadrature rules we present to calculate the non-zero entries to the sparse operators.

\noindent\secref{Computation}: We discuss computational issues, in particular, how to realise the results of the preceding sections in practice. We present a method for explicitly deriving the recurrence coefficients of the non-classical 1D OPs we detail in \secref{OPs}. We derive a quadrature rule on the disk-slice that can be used to expand a function in the OP basis up to a given order. Further, we implement function evaluation using the coefficients of the expansion of a given function using the Clenshaw algorithm.

\noindent\secref{Examples}: We demonstrate the proposed technique for solving Poisson, Helmholtz, and Biharmonic equations on the disk-slice.

\noindent\appref{PFEM}: We use the procedure to construct sparse $p$-finite element methods. This lays the groundwork for a future $hp$-finite element method in a disk, where the elements capture the circular geometry precisely.

\noindent\appref{HalfDisk}: We discuss extension to the special case of end-disk-slices (e.g., half disks).

\noindent\appref{Trapezium}: We discuss extension to trapezia.

@@

In this section we outline the construction and some basic properties of $\hdopnkabcd(x,y)$. The symmetry in the weight allows us to express the polynomials in terms of 1D OPs, and deduce certain properties such as recurrence relationships.

@@

\begin{proposition}[{\cite[p55--56]{dunkl2014orthogonal}}]\label{prop:construction}
Let \(w_1 : (\alpha,\beta) \: \to \R\), \(w_2 : (\gamma,\delta) \: \to \R\) be weight functions with \(\alpha,\beta,\gamma,\delta \in \R\), and let \(\rho \: : \: (\alpha,\beta) \: \to (0,\infty)\) be such that either \condref{trap} or \condref{disk} with \(w_2\) being an even function hold.
$\forall$, $n = 0,1,2,\dots, $ let $\{p_{n,k}\}$ be polynomials orthogonal with respect to the weight $\rho(x)^{2k+1} w_1(x)$ where $0 \le k \le n$, and $\{q_{n}\}$ be polynomials orthogonal with respect to the weight $w_2(x)$. Then the 2D polynomials defined on $\Omega$
$$
\hdopnk(x,y) := p_{n-k,k}(x) \: \rho(x)^k \: q_k\fpr(\frac{y}{\rho(x)}) \qquad\hbox{for} \qquad 0 \le k \le n, \: n = 0,1,2,\dots
$$
are orthogonal polynomials with respect to the weight \(W(x,y) := w_1(x) \: w_2\fpr(\frac{y}{\rho(x)}) \) on $\Omega$. 
\end{proposition}

For disk slices and trapeziums, we specialise Proposition \ref{prop:construction} in the following definition. First we introduce notation for two families of univariate OPs.
\begin{definition}\label{def:OPconstruction}
Let $\genjacw^{(a,b,c)}(x)$ and $\jacw^{(a,b)}(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\begin{align*}
\begin{cases}
\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
\jacw^{(a,b)}(x) &:= (\delta-x)^a \: (x - \gamma)^b
\end{cases}
\end{align*}
and define the associated inner products by:
\begin{align}
	\ip< p, \: q >_{\genjacw^{(a,b,c)}} &:= \frac{1}{\normgenjac^{(a,b,c)}} \: \int_\alpha^\beta p(x) \: q(x) \: \genjacw^{(a,b,c)}(x) \: \D x \label{eqn:ipgenjac} \\
	\ip<p, q>_{\jacw^{(a,b)}} &:= \frac{1}{\normjac^{(a,b)}} \: \int_\gamma^\delta p(y) \: q(y) \: \jacw^{(a,b)}(y)\: \D y \label{eqn:ipjac}
\end{align}
where
\begin{align}
	\normgenjac^{(a,b,c)} := \int_\alpha^\beta \: \genjacw^{(a,b,c)}(x) \: \D x, \quad \normjac^{(a,b)} := \int_\gamma^\delta \: \jacw^{(a,b)}(y) \: \D y. \label{eqn:ipnormalisations}
\end{align}
Denote the three-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjac_n^{(a,b,c)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipgenjac}), and the two-parameter family of orthonormal polynomials on $[\gamma,\delta]$ by $\{\jac_n^{(a,b)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipjac}).
\end{definition}

\begin{definition}\label{def:constuction}
Define the four-parameter 2D orthogonal polynomials via:
\begin{align*}
	\hdopnk^{(a,b,c,d)}(x,y) := \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: \jac_k^{(d,c)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align*}
\end{definition}

$\{\hdopnk^{(a,b,c,d)}\}$ are orthogonal with respect to the weight
\begin{align*}
	W^{(a,b,c,d)}(x,y) := \genjacw^{(a,b,c+d)}(x) \: \jacw^{(d,c)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega,
\end{align*}
assuming that either \condref{trap} or \condref{disk} with  \(\jacw^{(a,b)}\) being an even function (i.e. $a = b$, and we can hence denote the weight as $\jacw^{(a)}(x) = w^{(a,a)}_P(x) =  (\delta-x^2)^a$) hold.
That is,
\begin{align*}
	\ip< \hdopnk^{(a,b,c,d)}, \: \hdopmj^{(a,b,c,d)} >_{W^{(a,b,c,d)}} &= \normgenjac^{(a,b,c+d+2k+1)} \: \normjac^{(d,c)} \: \delta_{n,m} \: \delta_{k,j},
\end{align*}
where for $f, g : \Omega \to \R$ the inner product is defined as 
\begin{align*}
	\ip< f, \: g >_{W^{(a,b,c,d)}} := \iint_\Omega \: f(x,y) \: g(x,y) \: W^{(a,b,c,d)}(x,y) \: \D y \: \D x.
\end{align*}

We can see that they are indeed orthogonal using the change of variable $t = \frac{y}{\rho(x)}$, for the following normalisation:
\begin{align}
	&\ip< \hdopnkabcd, \: \hdopmjabcd >_\Wabcd \\
	&\quad \quad \quad =  \iint_\Omega \: \Big[ \genjacnmk^{(a,b,c+d+2k+1)}(x) \: \genjacmmj^{(a,b,c+d+2j+1)}(x) \: \rho(x)^{k+j}  \nonumber \\
		&\quad \quad \quad \quad \quad \quad \cdot \: \jac_k^{(d,c)}\fpr(\frac{y}{\rho(x)}) \: \jac_j^{(d,c)}\fpr(\frac{y}{\rho(x)})  \: \Wabcd(x,y) \Big] \: \D y \: \D x \nonumber \\
	&\quad \quad \quad = \Big( \int_\alpha^\beta \: {\genjacnmk^{(a,b,c+d+2k+1)}}(x) \: {\genjacmmj^{(a,b,c+d+2j+1)}}(x) \: \genjacw^{(a,b,c+d+k+j+1)}(x) \: \D x \Big) \nonumber \\
	&\quad \quad \quad \quad \quad \quad \cdot \: \Big( \int_\gamma^\delta \: {\jac_k^{(d,c)}}(t) \:  {\jac_j^{(d,c)}}(t) \: \jacw^{(d,)}(t) \: \D t \Big) \nonumber \\
	&\quad \quad \quad = \normjac^{(d,c)} \: \delta_{k,j} \: \int_\alpha^\beta \: {\genjacnmk^{(a,b,c+d+2k+1)}}(x) \: {\genjac_{m-k}^{(a,b,c+d+2k+1)}}(x) \: \genjacw^{(a,b,c+d+2k+1)}(x) \: \D x \nonumber \\ 
	&\quad \quad \quad = \normgenjac^{(a,b,c+d+2k+1)} \: \normjac^{(d,c)} \: \delta_{n,m} \: \delta_{k,j}. \label{eqn:normhdop}
\end{align}

>>>>

\chapter{Disk slices and trapeziums}

The contents of the previous chapter on spherical harmonics can allow us to use similar techniques for developing sparse spectral methods for solving linear partial differential equations on other domains. Before we can return to the 3D surface realm, let's first achieve a foundation in two-dimensions, by looking at a special class of geometries that includes disk slices and trapeziums.

More precisely, we will consider the solution of partial differential equations on the domain
\bseqn{
	\Omega := \bsset{(x,y) \in \R^2}{\alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)} 
}
where  either of the following conditions hold:

\begin{condition}\label{cond:trap}
	$\rho$ is a degree 1 polynomial.
\end{condition}
\begin{condition}\label{cond:disk}
	$\rho$ is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\).
\end{condition}

For simplicity of presentation we focus on the disk-slice in this chapter, where $\rho(x) = \sqrt{1-x^2}$, $(\alpha,\beta) \subset (0, 1)$, and $(\gamma, \delta)  = (-1,1)$. However, we will discuss an extension to other geometries, including the half-disk and trapeziums.

We show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that  generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations \cite{olver2013fast} and its analogue on the disk \cite{vasil2016tensor} and triangle \cite{olver2018recurrence,olver2019triangle}.  On the disk-slice the family of weights we consider are of the form
\bseqn{
	\Wabc(x,y) = (\beta - x)^a \: (x - \alpha)^b \: (1-x^2-y^2)^c, \qfor \alpha \leq x \leq \beta,\quad -\rho(x) \leq y \leq \rho(x).
}
The corresponding OPs denoted $\hdopnkabc(x,y)$, where $n$ denotes the polynomial degree, and $0 \le k \le n$. We define these to be orthogonalised lexicographically, that is,
\bseqn{
	\hdopnkabc(x,y) = C_{n,k} x^{n-k} y^k + (\hbox{lower order terms})
}
where $C_{n,k} \neq 0$ and ``lower order terms''   includes degree $n$ polynomials of the form $x^{n-j} y^j$ where $j < k$. The precise normalization arises from their definition in terms of one-dimensional OPs in Definition~\ref{def:OPconstruction}.

@@

@@

We can mirror the approach for the spherical harmonics, in that we will outline the construction and some basic properties of our 2D OPs on $\Omega$ that we denote by $\hdopnkabcd(x,y)$. The symmetry in the weight allows us to express the polynomials in terms of 1D OPs, and deduce certain properties such as recurrence relationships.

@@

\begin{proposition}[{\cite[p55--56]{dunkl2014orthogonal}}]\label{prop:construction}
	Let $w_1 : (\alpha,\beta) \: \to \R$, $w_2 : (\gamma,\delta) \: \to \R$ be weight functions with $\alpha,\beta,\gamma,\delta \in \R$, and let $\rho \: : \: (\alpha,\beta) \: \to (0,\infty)$ be such that either \condref{trap} or \condref{disk} with $w_2$ being an even function hold.
$\forall$, $n = 0,1,2,\dots, $ let $\{p_{n,k}\}$ be polynomials orthogonal with respect to the weight $\rho(x)^{2k+1} w_1(x)$ where $0 \le k \le n$, and $\{q_{n}\}$ be polynomials orthogonal with respect to the weight $w_2(x)$. Then the 2D polynomials defined on $\Omega$
\bseqn{
	\hdopnk(x,y) := p_{n-k,k}(x) \: \rho(x)^k \: q_k\fpr(\frac{y}{\rho(x)}) \qquad\hbox{for} \qquad 0 \le k \le n, \: n = 0,1,2,\dots
}
are orthogonal polynomials with respect to the weight \(W(x,y) := w_1(x) \: w_2\fpr(\frac{y}{\rho(x)}) \) on $\Omega$. 
\end{proposition}

For disk slices and trapeziums, we specialise Proposition \ref{prop:construction} in the following definitions. First we introduce notation for two families of univariate OPs.
\begin{definition}\label{def:OPconstruction}
	Let $\genjacw^{(a,b,c)}(x)$ and $\jacw^{(a,b)}(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\bseqn{
	\begin{cases}
		\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
		\jacw^{(a,b)}(x) &:= (\delta-x)^a \: (x - \gamma)^b
	\end{cases}
}
and define the associated inner products by:
\bseqnnumber{
	\ip< p, \: q >_{\genjacw^{(a,b,c)}} &:= \frac{1}{\normgenjac^{(a,b,c)}} \: \int_\alpha^\beta p(x) \: q(x) \: \genjacw^{(a,b,c)}(x) \: \D x \label{eqn:ipgenjac} \\
	\ip<p, q>_{\jacw^{(a,b)}} &:= \frac{1}{\normjac^{(a,b)}} \: \int_\gamma^\delta p(y) \: q(y) \: \jacw^{(a,b)}(y)\: \D y \label{eqn:ipjac}
}
where
\bseqnnumber{
	\normgenjac^{(a,b,c)} := \int_\alpha^\beta \: \genjacw^{(a,b,c)}(x) \: \D x, \quad \normjac^{(a,b)} := \int_\gamma^\delta \: \jacw^{(a,b)}(y) \: \D y. \label{eqn:ipnormalisations}
}
Denote the three-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjac_n^{(a,b,c)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipgenjac}), and the two-parameter family of orthonormal polynomials on $[\gamma,\delta]$ by $\{\jac_n^{(a,b)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipjac}).
\end{definition}

We can now write down our 2D OP family.
\begin{definition}\label{def:constuction}
	Define the four-parameter 2D orthogonal polynomials via:
\bseqn{
	\hdopnk^{(a,b,c,d)}(x,y) := \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: \jac_k^{(d,c)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
}
\end{definition}

The comparison with the construction of the 3D spherical harmonics is evident here (the $\genjac$ OPs are in place of the Jacobi polynomials, and the $\jac$ OPs are in place of the complex exponentials, which are just Chebyshev polynomials). By defining our OPs in this way, we will be able to yield sparse and banded relations for the necessary operators involved in PDEs, just as we did before.

$\{\hdopnk^{(a,b,c,d)}\}$ are orthogonal with respect to the weight
\bseqn{
	W^{(a,b,c,d)}(x,y) := \genjacw^{(a,b,c+d)}(x) \: \jacw^{(d,c)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega,
}
assuming that either \condref{trap} or \condref{disk} with  \(\jacw^{(a,b)}\) being an even function (i.e. $a = b$, and we can hence denote the weight as $\jacw^{(a)}(x) = w^{(a,a)}_P(x) =  (\delta-x^2)^a$) hold.
That is,
\bseqn{
	\ip< \hdopnk^{(a,b,c,d)}, \: \hdopmj^{(a,b,c,d)} >_{W^{(a,b,c,d)}} &= \normgenjac^{(a,b,c+d+2k+1)} \: \normjac^{(d,c)} \: \delta_{n,m} \: \delta_{k,j},
}
where for $f, g : \Omega \to \R$ the inner product is defined as 
\bseqn{
	\ip< f, \: g >_{W^{(a,b,c,d)}} := \iint_\Omega \: f(x,y) \: g(x,y) \: W^{(a,b,c,d)}(x,y) \: \D y \: \D x.
}

We can see that they are indeed orthogonal using the change of variable $t = \frac{y}{\rho(x)}$, for the following normalisation:
\bseqnnumber{
	&\ip< \hdopnkabcd, \: \hdopmjabcd >_\Wabcd \\
	&\quad \quad \quad =  \iint_\Omega \: \Big[ \genjacnmk^{(a,b,c+d+2k+1)}(x) \: \genjacmmj^{(a,b,c+d+2j+1)}(x) \: \rho(x)^{k+j}  \nonumber \\
		&\quad \quad \quad \quad \quad \quad \cdot \: \jac_k^{(d,c)}\fpr(\frac{y}{\rho(x)}) \: \jac_j^{(d,c)}\fpr(\frac{y}{\rho(x)})  \: \Wabcd(x,y) \Big] \: \D y \: \D x \nonumber \\
	&\quad \quad \quad = \Big( \int_\alpha^\beta \: {\genjacnmk^{(a,b,c+d+2k+1)}}(x) \: {\genjacmmj^{(a,b,c+d+2j+1)}}(x) \: \genjacw^{(a,b,c+d+k+j+1)}(x) \: \D x \Big) \nonumber \\
	&\quad \quad \quad \quad \quad \quad \cdot \: \Big( \int_\gamma^\delta \: {\jac_k^{(d,c)}}(t) \:  {\jac_j^{(d,c)}}(t) \: \jacw^{(d,)}(t) \: \D t \Big) \nonumber \\
	&\quad \quad \quad = \normjac^{(d,c)} \: \delta_{k,j} \: \int_\alpha^\beta \: {\genjacnmk^{(a,b,c+d+2k+1)}}(x) \: {\genjac_{m-k}^{(a,b,c+d+2k+1)}}(x) \: \genjacw^{(a,b,c+d+2k+1)}(x) \: \D x \nonumber \\ 
	&\quad \quad \quad = \normgenjac^{(a,b,c+d+2k+1)} \: \normjac^{(d,c)} \: \delta_{n,m} \: \delta_{k,j}. \label{eqn:normhdop}
}

For the disk-slice, the weight $\Wabc(x,y) = (\beta - x)^a \: (x - \alpha)^b \: (1-x^2-y^2)^c$ results from setting:
\bseqn{
\begin{cases}
	(\alpha,\beta) &\subset (0,1) \\
	(\gamma,\delta) &:= (-1,1) \\
	\rho(x) &:= (1-x^2)^{\half}
\end{cases}
}
so that
\bseqn{
\begin{cases}
	\genjacw^{(a,b,c)}(x) :=  (\beta - x)^a \: (x - \alpha)^b \: \rho(x)^{c} \\
	\jacw^{(c)}(x) := (1-x)^c \: (1+x)^c = (1-x^2)^c.
\end{cases}
}
Note here we can simply remove the need for including a fourth parameter $d$. The 2D OPs orthogonal with respect to the weight above on the disk-slice $\Omega$ are then given by:
\bseqnnumber{\label{eq:diskpolys}
	\hdopnkabc(x,y) := \genjacnmk^{(a, b, 2c+2k+1)}(x) \: \rho(x)^k \: \jac_k^{(c,c)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega
}
In this case the weight $\jacw(x)$ is an ultraspherical weight, and the corresponding OPs are the normalized Jacobi polynomials $\{\jac_n^{(b, b)}\}$, while the weight $w_R(x)$ is non-classical (it is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}).
